---
layout: post
title: "Spatial Relations and Besag-York-Mollie Application"
author: "Rasim M Musal"
date: "2023-12-24"
output:
  html_document:
   theme: darkly
   highlight: espresso
   toc: true
   keep_md: yes
   toc_float: true
   toc_collapsed: false
   toc_depth: 3
   number_sections: true
   usemathjax: true
bibliography: bibliography.bib 
tags: [Besag-York-Mollie, cmdstan, STAN, R]
always_allow_html: true
editor_options: 
  markdown: 
    wrap: 72
---

```{=html}
<script type="text/x-mathjax-config">
MathJax.Hub.Config({
  TeX: { 
      equationNumbers: {
 
            autoNumber: "all",
            formatNumber: function (n) {return +n}
      } 
  }
});
</script>
```
```{r setup, include=FALSE,warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE,warning=FALSE}
# A function for generating captions and cross-references

fig <- local({
    i <- 0
    list(
        cap=function(refName, text, center=FALSE, col="black", inline=FALSE) {
            i <<- i + 1
            ref[[refName]] <<- i
            css_ctr <- ""
            if (center) css_ctr <- "text-align:center; display:inline-block; width:100%;"
            cap_txt <- paste0("<span style=\"color:", col, "; ", css_ctr, "\">Figure ", i, ": ", text , "</span>")
            anchor <- paste0("<a name=\"", refName, "\"></a>")
            if (inline) {
                paste0(anchor, cap_txt)    
            } else {
                list(anchor=anchor, cap_txt=cap_txt)
            }
        },
        
        ref=function(refName, link=FALSE, checkRef=TRUE) {
            
            ## This function puts in a cross reference to a caption. You refer to the
            ## caption with the refName that was passed to fig$cap() (not the code chunk name).
            ## The cross reference can be hyperlinked.
            
            if (checkRef && !refName %in% names(ref)) stop(paste0("fig$ref() error: ", refName, " not found"))
            if (link) {
                paste0("<A HREF=\"#", refName, "\">Figure ", ref[[refName]], "</A>")
            } else {
                paste0("Figure ", ref[[refName]])
            }
        },
        
        ref_all=function(){
            ## For debugging
            ref
        })
})
library(knitr)
knit_hooks$set(plot = function(x, options) {
    sty <- ""
    if (options$fig.align == 'default') {
        sty <- ""
    } else {
        sty <- paste0(" style=\"text-align:", options$fig.align, ";\"")
    }
    
    if (is.list(options$fig.cap)) {
        ## options$fig.cap is a list returned by the function fig$cap()
        str_caption <- options$fig.cap$cap_txt
        str_anchr <- options$fig.cap$anchor
    } else {
        ## options$fig.cap is a character object (hard coded, no anchor)
        str_caption <- options$fig.cap
        str_anchr <- ""
    }
    
    paste('<figure', sty, '>', str_anchr, '<img src="',
        opts_knit$get('base.url'), paste(x, collapse = '.'),
        '"><figcaption>', str_caption, '</figcaption></figure>',
        sep = '')
    
})

rmdFn <- knitr::current_input()  # filename of input document

## Read lines and close connection
rmdCon <- file(rmdFn, open = "r")
rmdLines <- readLines(rmdCon)
close(rmdCon)

## Pull out all occurences of at least one back tick, followed 
## by any number of characters, followed by fig$cap (all on one line)
figscap_idx <- grep("`+(.*)fig\\$cap", rmdLines)
rmdLines <- rmdLines[figscap_idx]

## Get rid of everything up until the start of the caption label
## This presumes the caption label is the first argument of fig$cap()
## E.g., fig.cap = fig$cap("my_label", ...)
rmdLinesSansPre <- sub("(.*)fig\\$cap(.*?)[\"']", "", rmdLines)

## Identify everything up until the first quote
match_data <- regexpr("(.*?)[\"']", rmdLinesSansPre)

## Reduce the length by one, because we're not interested in the final quote
attr(match_data, "match.length") <- attr(match_data, "match.length") - 1

## Extract
fig_labels <- regmatches(rmdLinesSansPre, match_data, invert=FALSE)

if (length(fig_labels) > 0) {

    ## Test for duplicates
    if (anyDuplicated(fig_labels) > 0) stop("Duplicate caption labels detected")
    
    ## Create a named list of Figure numbers
    ref <- as.list(1:length(fig_labels))
    names(ref) <- fig_labels
}    
```

# Compiling and Running via cmdSTAN in UBUNTU
The first step in compiling our program is to navigate to the cmdstan file. 
```{bash warning=FALSE,message=FALSE,eval=FALSE}
cd /home/admin/cmdstan
```

We then have cmdstan turn our modelA1 into an executable program via the make command.

```{bash warning=FALSE,message=FALSE,eval=FALSE}
make /home/admin/Desktop/research/modelA1
```

Navigate to the folder where the modelA1.STAN has an executable file under the name modelA1.

```{bash warning=FALSE,message=FALSE,eval=FALSE}
cd /home/admin/Desktop/research/
```

While remaining in the same location, we instruct cmdstan to create 3 MCMC chains to run in parallel, do a 50,000 set of warmup simulations and sample 10,000 from them. The data is specified as '2023noICUcoviddata1.r' and the output files will be specified as mA11 with its extensions specified as _i leading to the file names mA11_1,mA11_2, and mA11_3. The execution of the model took approximately 9 hrs in hour workstation.

```{bash warning=FALSE,message=FALSE,eval=FALSE}
for i in {1..3}; do ./modelA1 sample num_warmup=50000
num_samples=10000 data file=./2023noICUcoviddata1.r 
output file output file=mA11_${i} &done&
```

# Reading the files and checking convergence

In this section we will read in the csv files and check for convergence. Note that the mA11_1,mA11_2 and mA11_3 do not have file extensions so we added them for convenience. The model created these large files and they will take a large RAM capacity in your R session.   

```{r, warning=FALSE,message=FALSE}
#Loading the libraries and specifying the options to use multiple cores as needed
options(mc.cores=parallel::detectCores(),auto_write = TRUE)
library(rstan)
library(xtable)
library(ggspatial)
library(ggplot2)
library(viridis)
setwd("C:/Users/rm84/Desktop/research/HMM/data")
load("workspacewithbasedata.RData")
#Specifying the location of the csv files and creating string objects

chain1='C:/Users/rm84/Documents/mA11_1.csv'
chain2='C:/Users/rm84/Documents/mA11_2.csv'
chain3='C:/Users/rm84/Documents/mA11_3.csv'
```

To check for convergence we read in the csv files and obtain summary information of the MCMC samples. There is a good discussion on why we should not focus only on Rhat^[https://statmodeling.stat.columbia.edu/2019/03/19/maybe-its-time-to-let-the-old-ways-die-or-we-broke-r-hat-so-now-we-have-to-fix-it/] however we know that if Rhat is over 1.01 we definitely do not have a well mixed model. We show here that there were no parameters with an Rhat over 1.01 and proceed to also highlight the similarity between the observed mortality data and the generated values.  

```{r, warning=FALSE,message=FALSE}
csvfiles=c(chain1,chain2,chain3)
fit=read_stan_csv(csvfiles)
fitsummary=summary(fit)$summary

Rhats=fitsummary[,"Rhat"]

sum(Rhats>1.01,na.rm=TRUE)
Rhats[Rhats>1.01]

```

# Converting the samples to R object

We proceed to extract all the samples into the object listofdraws, remove the stan fit object, use the garbage collection gc() function to release RAM. Attaching the listofdraws allow us to refer to the parameters directly. 

```{r, warning=FALSE,message=FALSE}
listofdraws=as.data.frame(rstan::extract(fit))
rm(fit)
gc()
#rho 
attach(listofdraws)

```

# Obtaining Root Mean Square Error and DIC4

The list of lists, object named 'cadata', contain all the data that our model have used. We will first obtain the root mean square error in rMSE object. We are going to utilize the $DIC_4$ framework from @celeux06 to calculate the fit our model to compare to others later. The result is contained in the object ll which we report.     

```{r, warning=FALSE,message=FALSE}
source("C:/Users/rm84/Desktop/research/HMM/data/R2023noICUcoviddata1.r")
START=1
END=77
ITER=END-START+1
cadata$START=START
cadata$END=END

meansoflambda=array(dim=c(ITER,length=N))
for(t in START:END){
  for(n in 1:N){
    meansoflambda[t-START+1,n]=mean(eval(parse(text=(paste0("alpha.",t-START+1,".",1,".",n)))))
  }}

MSEarray=array(dim=c(ITER))
for(t in START:END){
  MSEarray[t-START+1]=mean((cadata$y[t,]-meansoflambda[t-START+1,])^2)
}
MSE=mean(MSEarray)
rMSE=sqrt(MSE)

rMSE

l=length(alpha.1.1.1)
ll1=array(dim=c(l,(END-START+1),N))
for(t in START:END){
  for(n in 1:N){
    ll1[,t,n]=dpois(y[t,n],lambda=eval(parse(text=(paste0("alpha.",t-START+1,".",1,".",n)))),log=TRUE)
  }}

ll2=matrix(nrow=(END-START+1),ncol=N)
for(t in START:END){
  for(n in 1:N){
    ll2[t,n]=dpois(y[t,n],lambda=mean(eval(parse(text=(paste0("alpha.",t-START+1,".",1,".",n))))),log=TRUE)
  }}


ll=-4*mean(rowSums(ll1))+2*sum(ll2)

ll

```

# Analysis of parameters
This section will discuss the findings associated with modelA1. We will first discuss the fixed parameters of interest associated with the covariates. We will follow that with a discussion of coefficients of Vaccinations through $t=1 \ldots t=77$. We proceed to discuss $\phi$ and $\theta$ terms. 

## Parameters of Fixed Covariates

```{r, warning=FALSE,message=FALSE}


summary1=xtable(summary((cbind.data.frame(beta_pov,
beta_inc,beta_dens,beta_gini,beta_white,beta_age,beta_sex))))

s1=(as.data.frame(cbind(beta_pov,beta_inc,beta_dens,beta_gini,
beta_white,beta_age,beta_sex)))

m1=as.data.frame(matrix(data=s1,nrow=6,ncol=7))

  m1[1,]=round(sapply(s1,min),3)
  m1[2,]=round(apply(s1[1:7],2,quantile,probs=0.025),3)
  m1[3,]=round(sapply(s1,median),3)
  m1[4,]=round(colMeans(s1),3)
  m1[5,]=round(apply(s1[1:7],2,quantile,probs=0.975),3)
  m1[6,]=round(sapply(s1,max),3)
  row.names(m1) = c("Min","2.5th Perc.","Med.","Mean","97.5th Perc.","Max")
knitr::kable(m1,
col.names = c("$\\beta_{Pov}$","$\\beta_{Inc}$","$\\beta_{Dens}$","$\\beta_{gini}$","$\\beta_{white}$","$\\beta_{age}$","$\\beta_{sex}$"),
escape=FALSE)


```
As can be seen from the table's $95\%$ credibility intervals increased poverty, density, percentage of people who are white, median age leads to increased mortality risk. On the other hand increased income and males to females ratio in the county decreases mortality risk. When we control for all the covariates, Gini covariate coefficient contains the value 0 in the $95\%$ credibility interval.

## Parameters of Vaccination Coefficients

As can be demonstrated from the plot in this section we can clearly see that in most of 2021 the vaccinations decreased mortality risk in the counties of california. The only biweek where vaccinations in the counties of California had a positive relationship to mortality risk is in biweek 62 which is in 2022. For all of the rest of 2022 the credibility interval contains 0. This acquires more attention and could be due to the vaccinations losing their effectiveness and most of the population having some degree of protection. As both the disease and the vaccinations are evolving we would need to pay close attention to the dynamic nature of these effects. In fact we argue that we should be looking at more changes in the fixed covariates' effect through the biweeks.     

```{r, warning=FALSE,message=FALSE}
library(ggpubr)
library(dplyr)
library(gridExtra)

V=cadata$V
qhigh=array(dim=c(T-V+1))
qlow=array(dim=c(T-V+1))
beta_vac=as.data.frame(matrix(nrow=(T-V+1)*l,ncol=2))

for(i in 1:(T-V+1)){
  beta_vac[((i-1)*(l)+1):((i)*l),1]=eval(parse(text=(paste0("beta_vac.",i))))
  beta_vac[((i-1)*(l)+1):((i)*l),2]=i
  #qhigh[i]=quantile(eval(parse(text=(paste0("beta_vac.",i)))),0.975)
  #qlow[i]=quantile(eval(parse(text=(paste0("beta_vac.",i)))),0.025)
}
names(beta_vac)=c("beta_vac","Biweeks")

qhigh_data <- beta_vac %>%
  group_by(Biweeks) %>%
  summarise(qhigh = quantile(beta_vac,0.975))

qlow_data <- beta_vac %>%
  group_by(Biweeks) %>%
  summarise(qlow = quantile(beta_vac,0.025))

  beta_vacplot=ggplot(data = beta_vac,aes(x= Biweeks, y = beta_vac,group=Biweeks))+
  geom_boxplot(outlier.shape = NA)+scale_y_continuous(name=bquote(beta[vac]))+ggtitle(bquote(beta[vac] ~ "from t=1...77"))+theme(plot.title = element_text(hjust = 0.5))+
  geom_hline(yintercept=0.0, linetype="solid", color = "red") +
  scale_x_continuous(name="Time",labels=c('25','29','34','39','44','49','54','59','64','69','74','77'),breaks=c(1,5,10,15,20,25,30,35,40,45,50,55))+
    stat_summary(data = qhigh_data, 
                 aes(x = Biweeks, y = qhigh),
                 geom = "point", 
                 size = 1, 
                 color = "red")+
    stat_summary(data = qlow_data, 
                 aes(x = Biweeks, y = qlow),
                 geom = "point", 
                 size = 1, 
                 color = "red")
beta_vacplot
#Number of biweeks where 0 is above the Credibility Interval  
num_biweeks_CI_neg<-sum(0<qlow_data$qlow)   
num_biweeks_CI_neg
#What were the biweeks where 0 is above the Credibility Interval 
which(0<qlow_data$qlow)+V
#Number of biweeks where 0 is below the Credibility Interval
num_biweeks_CI_pos<-sum(0>qhigh_data$qhigh)
num_biweeks_CI_pos
#What were the biweeks where 0 is below the Credibility Interval
which(0>qhigh_data$qhigh)+V


```

## $\theta$, Random Effect Coefficients of Counties

We proceed to illustrate the random effects of counties. It is not surprising that for most counties, $\theta_{it}$ which can be referred to as the county performance varies across the biweeks per county. In the script below, the R object thetaXX is going to contain the MCMC samples from various counties. Per the rest of the project we are going to look in detail to 6 Californian counties. The XX values of the components are given in parenthesis after county names. Los Angeles (19), San Diego (37), San Francisco (38), Kern (15), Lassen (18) and San Bernardino (36).  We are following the random effects for 77 biweeks after we take into account the listed covariates as well as areal effects of the counties surrounding them. The posterior means and credibility intervals (CI) of this component are of special importance to policy makers as relatively high/low values should have them look into the root causes behind low/high increase in relative mortality risk.

In order to visualize the random effects via ggplot we will be creating data frame objects with M rows and 2 columns for each of the before mentioned counties. M is the number of MCMC samples multiplied by the biweeks. The first column is going to contain the MCMC values and the second column the biweek value. 

Each of the counties will have a ggplot associated with it that visualizes the boxplot of the MCMC samples for each biweek as well as the Credibility Interval (CI).    

```{r, warning=FALSE,message=FALSE}
library(ggpubr)
library(gridExtra)
library(ggplot2)
library(tidyverse)

cnames<-read.table("C:/Users/rm84/Desktop/research/HMM/data/countynames.txt",sep="\t",header=TRUE)[,1]

START=1
END=77
ITER=END-START+1

l=length(theta.1.15)

M=ITER*l


theta15=as.data.frame(matrix(nrow=M,ncol=2))


n=15
for(t in 1:ITER){
  theta15[((t-1)*(l)+1):((t)*l),1]=eval(parse(text=(paste0("theta.",t,".",n))))
  theta15[((t-1)*(l)+1):((t)*l),2]=t

}

names(theta15)=c("theta","Biweeks")


theta18=as.data.frame(matrix(nrow=M,ncol=2))
n=18
for(t in 1:ITER){
  theta18[((t-1)*(l)+1):((t)*l),1]=eval(parse(text=(paste0("theta.",t,".",n))))
  theta18[((t-1)*(l)+1):((t)*l),2]=t
}

names(theta18)=c("theta","Biweeks")

theta36=as.data.frame(matrix(nrow=M,ncol=2))
n=36
for(t in 1:ITER){
  theta36[((t-1)*(l)+1):((t)*l),1]=eval(parse(text=(paste0("theta.",t,".",n))))
  theta36[((t-1)*(l)+1):((t)*l),2]=t
}

names(theta36)=c("theta","Biweeks")


#################################################
theta19=as.data.frame(matrix(nrow=M,ncol=2))
n=19
for(t in 1:ITER){
  theta19[((t-1)*(l)+1):((t)*l),1]=eval(parse(text=(paste0("theta.",t,".",n))))
  theta19[((t-1)*(l)+1):((t)*l),2]=t
}

names(theta19)=c("theta","Biweeks")

theta37=as.data.frame(matrix(nrow=M,ncol=2))
n=37
for(t in 1:ITER){
  theta37[((t-1)*(l)+1):((t)*l),1]=eval(parse(text=(paste0("theta.",t,".",n))))
  theta37[((t-1)*(l)+1):((t)*l),2]=t
}

names(theta37)=c("theta","Biweeks")

theta38=as.data.frame(matrix(nrow=M,ncol=2))
n=38
for(t in 1:ITER){
  theta38[((t-1)*(l)+1):((t)*l),1]=eval(parse(text=(paste0("theta.",t,".",n))))
  theta38[((t-1)*(l)+1):((t)*l),2]=t
}

names(theta38)=c("theta","Biweeks")

############################################################################################

###Kern
qhigh_datatheta15 <- theta15 %>%
  group_by(Biweeks) %>%
  summarise(qhigh_theta15 = quantile(theta,0.975))

qlow_datatheta15 <- theta15 %>%
  group_by(Biweeks) %>%
  summarise(qlow_theta15 = quantile(theta,0.025))

###Lassen 
qhigh_datatheta18 <- theta18 %>%
  group_by(Biweeks) %>%
  summarise(qhigh_theta18 = quantile(theta,0.975))

qlow_datatheta18 <- theta18 %>%
  group_by(Biweeks) %>%
  summarise(qlow_theta18 = quantile(theta,0.025))

###Los Angeles
qhigh_datatheta19 <- theta19 %>%
  group_by(Biweeks) %>%
  summarise(qhigh_theta19 = quantile(theta,0.975))

qlow_datatheta19 <- theta19 %>%
  group_by(Biweeks) %>%
  summarise(qlow_theta19 = quantile(theta,0.025))

### San Bernardino
qhigh_datatheta36 <- theta36 %>%
  group_by(Biweeks) %>%
  summarise(qhigh_theta36 = quantile(theta,0.975))

qlow_datatheta36 <- theta36 %>%
  group_by(Biweeks) %>%
  summarise(qlow_theta36 = quantile(theta,0.025))
### San Diego
qhigh_datatheta37 <- theta37 %>%
  group_by(Biweeks) %>%
  summarise(qhigh_theta37 = quantile(theta,0.975))

qlow_datatheta37 <- theta37 %>%
  group_by(Biweeks) %>%
  summarise(qlow_theta37 = quantile(theta,0.025))
##### San Francisco
qhigh_datatheta38 <- theta38 %>%
  group_by(Biweeks) %>%
  summarise(qhigh_theta38 = quantile(theta,0.975))

qlow_datatheta38 <- theta38 %>%
  group_by(Biweeks) %>%
  summarise(qlow_theta38 = quantile(theta,0.025))
####  

la=ggplot(data = theta19,aes(x= Biweeks, y = theta,group=Biweeks))+
  geom_boxplot(outlier.shape = NA)+ggtitle(cnames[19])+theme(plot.title = element_text(hjust = 0.5))+
  geom_hline(yintercept=0, linetype="solid", color = "red")+scale_y_continuous(name=bquote(theta[19~",t"]))+
  scale_x_continuous(name="Time",labels=c('1','25','51','77'),breaks=c(1,25,51,77))+stat_summary(data = qhigh_datatheta19, 
                 aes(x = Biweeks, y = qhigh_theta19),
                 geom = "point", 
                 size = 1, 
                 color = "red")+
    stat_summary(data = qlow_datatheta19, 
                 aes(x = Biweeks, y = qlow_theta19),
                 geom = "point", 
                 size = 1, 
                 color = "red")

sd=ggplot(data = theta37,aes(x= Biweeks, y = theta,group=Biweeks))+
  scale_y_continuous(name=bquote(theta))+
  geom_boxplot(outlier.shape = NA)+ggtitle(cnames[37])+theme(plot.title = element_text(hjust = 0.5))+
  geom_hline(yintercept=0, linetype="solid", color = "red")+scale_y_continuous(name=bquote(theta[37~",t"]))+
  scale_x_continuous(name="Time",labels=c('1','25','51','77'),breaks=c(1,25,51,77))+stat_summary(data = qhigh_datatheta37, 
                 aes(x = Biweeks, y = qhigh_theta37),
                 geom = "point", 
                 size = 1, 
                 color = "red")+
    stat_summary(data = qlow_datatheta37, 
                 aes(x = Biweeks, y = qlow_theta37),
                 geom = "point", 
                 size = 1, 
                 color = "red")

sf=ggplot(data = theta38,aes(x= Biweeks, y = theta,group=Biweeks))+
  scale_y_continuous(name=bquote(theta))+
  geom_boxplot(outlier.shape = NA)+ggtitle(cnames[38])+theme(plot.title = element_text(hjust = 0.5))+
  geom_hline(yintercept=0, linetype="solid", color = "red")+scale_y_continuous(name=bquote(theta[38~",t"]))+
  scale_x_continuous(name="Time",labels=c('1','25','51','77'),breaks=c(1,25,51,77))+stat_summary(data = qhigh_datatheta38,                 aes(x = Biweeks, y = qhigh_theta38),
                 geom = "point", 
                 size = 1, 
                 color = "red")+
    stat_summary(data = qlow_datatheta38, 
                 aes(x = Biweeks, y = qlow_theta38),
                 geom = "point", 
                 size = 1, 
                 color = "red")

#######################################################################################

Lassen=ggplot(data = theta18,aes(x= Biweeks, y = theta,group=Biweeks))+
  geom_boxplot(outlier.shape = NA)+ggtitle(cnames[18])+theme(plot.title = element_text(hjust = 0.5))+
  geom_hline(yintercept=0, linetype="solid", color = "red")+ylim(-5,5)+scale_y_continuous(name=bquote(theta[18~",t"]))+
  scale_x_continuous(name="Time",labels=c('1','25','51','77'),breaks=c(1,25,51,77))+stat_summary(data = qhigh_datatheta18, 
                 aes(x = Biweeks, y = qhigh_theta18),
                 geom = "point", 
                 size = 1, 
                 color = "red")+
    stat_summary(data = qlow_datatheta18, 
                 aes(x = Biweeks, y = qlow_theta18),
                 geom = "point", 
                 size = 1, 
                 color = "red")

#############################################################
Kern=ggplot(data = theta15,aes(x= Biweeks, y = theta,group=Biweeks))+
  geom_boxplot(outlier.shape = NA)+ggtitle(cnames[15])+theme(plot.title = element_text(hjust = 0.5))+
  geom_hline(yintercept=0, linetype="solid", color = "red")+ylim(-5,5)+
  scale_y_continuous(name=bquote(theta[15~",t"]))+
  scale_x_continuous(name="Time",labels=c('1','25','51','77'),breaks=c(1,25,51,77))+stat_summary(data = qhigh_datatheta15,     aes(x = Biweeks, y = qhigh_theta15),
                 geom = "point", 
                 size = 1, 
                 color = "red")+
    stat_summary(data = qlow_datatheta15, 
                 aes(x = Biweeks, y = qlow_theta15),
                 geom = "point", 
                 size = 1, 
                 color = "red")

#############################################################
sbern=ggplot(data = theta36,aes(x= Biweeks, y = theta,group=Biweeks))+
  geom_boxplot(outlier.shape = NA)+ggtitle(cnames[36])+theme(plot.title = element_text(hjust = 0.5))+
  geom_hline(yintercept=0, linetype="solid", color = "red")+ylim(-5,5)+
  scale_y_continuous(name=bquote(theta[36~",t"]))+
  scale_x_continuous(name="Time",labels=c('1','25','51','77'),breaks=c(1,25,51,77))+stat_summary(data = qhigh_datatheta36, 
                 aes(x = Biweeks, y = qhigh_theta36),
                 geom = "point", 
                 size = 1, 
                 color = "red")+
    stat_summary(data = qlow_datatheta36, 
                 aes(x = Biweeks, y = qlow_theta36),
                 geom = "point", 
                 size = 1, 
                 color = "red")




```

Out of the 6 counties we are looking into, the first county we will investigate is San Bernardino.

Looking at the boxplot of $\theta$ vs biweeks of San Bernardino; 

```{r, echo=FALSE,warning=FALSE,message=FALSE}

sbern

```

we see patterns of some note indicated by CI. In 2020 there is not a lot to note except the few biweeks in the middle of the year that suggests the reduction in mortality risk, which unfortunately turns to an increase in the last biweeks. For great majority of 2021 most of the credibility intervals contain the value 0. In the initial months of year 2022 that increased mortality risk which reverses direction in the latter moths. We can quantify the performance of the county by comparing the lower and upper bound values of the CI to 0. The lower (upper) bound value of CI being higher than 0 indicates higher (lower) relative mortality risk.   

Using the formulation 0 < 95% Lower Bound CI and summing up the number of biweeks we find that `r sum(0<qlow_datatheta36[,2])` of the times San Bernardino had higher relative risk. To find the number of biweeks with lower mortality risk we replace the 95% Lower Bound CI with 95% Upper Bound CI and change '>' to '<' which results in `r sum(0>qhigh_datatheta36[,2])` number of biweeks.  Policy makers should look at these across time and investigate the reason behind this performance specific to the county, after accounting covariates and areal effects.   


```{r, warning=FALSE,message=FALSE}
SanBernOutliers<-as.data.frame(matrix(nrow=2,ncol=3))

SanBernOutliers[1,1]<-sum(0>qhigh_datatheta36[1:25,2])
SanBernOutliers[1,2]<-sum(0>qhigh_datatheta36[26:51,2])
SanBernOutliers[1,3]<-sum(0>qhigh_datatheta36[52:77,2])

SanBernOutliers[2,1]<-sum(0<qlow_datatheta36[1:25,2])
SanBernOutliers[2,2]<-sum(0<qlow_datatheta36[26:51,2])
SanBernOutliers[2,3]<-sum(0<qlow_datatheta36[52:77,2])

row.names(SanBernOutliers) = c("Lower Risk","Higher Risk")
  
kable(SanBernOutliers,align="c",
col.names = c("  2020  ", "  2021  ", "  2022  "),caption=" The number of biweeks that the CI did not contain 0 indicating higher or lower relative mortality risk in San Bernardino",
escape=FALSE)


```

There really is not much to infer about Lassen's $\theta$ component and there does not seem to be much learning of the parameters. 

```{r, echo=FALSE,warning=FALSE,message=FALSE}

Lassen

```

Kern follows a similar pattern to San Bernardino. The boxplot vs biweeks will provide us with a varying pattern 

```{r, echo=FALSE,warning=FALSE,message=FALSE}

Kern

```

If we cross tabulate the lower/higher relative mortality risk effect by years:

```{r, echo=FALSE,warning=FALSE,message=FALSE}
KernOutliers<-as.data.frame(matrix(nrow=2,ncol=3))

KernOutliers[1,1]<-sum(0>qhigh_datatheta15[1:25,2])
KernOutliers[1,2]<-sum(0>qhigh_datatheta15[26:51,2])
KernOutliers[1,3]<-sum(0>qhigh_datatheta15[52:77,2])

KernOutliers[2,1]<-sum(0<qlow_datatheta15[1:25,2])
KernOutliers[2,2]<-sum(0<qlow_datatheta15[26:51,2])
KernOutliers[2,3]<-sum(0<qlow_datatheta15[52:77,2])

row.names(KernOutliers) = c("Lower Risk","Higher Risk")
  
knitr::kable(KernOutliers,align="c",
col.names = c("  2020  ", "  2021  ", "  2022  "),caption=("The number of biweeks that the CI did not contain 0 indicating higher or lower relative mortality risk in Kern"),
escape=FALSE)

```


Out of the 3 big cities we investigate Los Angeles first. There seems to be a larger number of biweeks in Los Angeles where the value 0 was not contained within the Credibility intervals. If we do a similar analysis to Kern we see; 

```{r, echo=FALSE,warning=FALSE,message=FALSE}
LAOutliers<-as.data.frame(matrix(nrow=2,ncol=3))

LAOutliers[1,1]<-sum(0>qhigh_datatheta19[1:25,2])
LAOutliers[1,2]<-sum(0>qhigh_datatheta19[26:51,2])
LAOutliers[1,3]<-sum(0>qhigh_datatheta19[52:77,2])

LAOutliers[2,1]<-sum(0<qlow_datatheta19[1:25,2])
LAOutliers[2,2]<-sum(0<qlow_datatheta19[26:51,2])
LAOutliers[2,3]<-sum(0<qlow_datatheta19[52:77,2])

row.names(LAOutliers) = c("Lower Risk","Higher Risk")
  
knitr::kable(LAOutliers,align="c",
col.names = c("2020", "2021", "2022"),caption=("The number of biweeks that the CI did not contain 0 indicating higher or lower relative mortality risk in Los Angeles"),
escape=FALSE)

```

San Diego had a more varied year in 2020 compared to 2021 and 2022. When we look at the same plot as the other counties in this section;

```{r,echo=FALSE,warning=FALSE,message=FALSE}

sd

```
it is straightforward from looking at the boxplots for the great majority of the time the county performed well even though there were certain weeks in 2020 and 2021 of poor performance. 

```{r, echo=FALSE,warning=FALSE,message=FALSE}
SDOutliers<-as.data.frame(matrix(nrow=2,ncol=3))

SDOutliers[1,1]<-sum(0>qhigh_datatheta37[1:25,2])
SDOutliers[1,2]<-sum(0>qhigh_datatheta37[26:51,2])
SDOutliers[1,3]<-sum(0>qhigh_datatheta37[52:77,2])

SDOutliers[2,1]<-sum(0<qlow_datatheta37[1:25,2])
SDOutliers[2,2]<-sum(0<qlow_datatheta37[26:51,2])
SDOutliers[2,3]<-sum(0<qlow_datatheta37[52:77,2])

row.names(SDOutliers) = c("Lower Risk","Higher Risk")
  
knitr::kable(SDOutliers,align="c",
col.names = c("2020", "2021", "2022"),caption=("The number of biweeks that the CI did not contain 0 indicating higher or lower relative mortality risk in San Diego"),
escape=FALSE)

```

We again provide the boxplots vs biweeks, this time for the county of San Francisco. 

```{r, echo=FALSE,warning=FALSE,message=FALSE}

sf

```

It is interesting to see that San Francisco as a county performed well until 2022. It would be of interest to understand the reasons behind the relative poor performance in that year.

```{r, echo=FALSE, warning=FALSE,message=FALSE}
SFOutliers<-as.data.frame(matrix(nrow=2,ncol=3))

SFOutliers[1,1]<-sum(0>qhigh_datatheta38[1:25,2])
SFOutliers[1,2]<-sum(0>qhigh_datatheta38[26:51,2])
SFOutliers[1,3]<-sum(0>qhigh_datatheta38[52:77,2])

SFOutliers[2,1]<-sum(0<qlow_datatheta38[1:25,2])
SFOutliers[2,2]<-sum(0<qlow_datatheta38[26:51,2])
SFOutliers[2,3]<-sum(0<qlow_datatheta38[52:77,2])

row.names(SFOutliers) = c("Lower Risk","Higher Risk")
  
knitr::kable(SFOutliers,align="c",
col.names = c("2020", "2021", "2022"),caption=("The number of biweeks that the CI did not contain 0 indicating higher or lower relative mortality risk in San Francisco"),
escape=FALSE)
```


## $\phi$ , The areal joint distribution 
When we look at the joint distribution of spatial contribution to mortality risk, out of the 

```{r, warning=FALSE,message=FALSE}

phis=as.data.frame(matrix(nrow=N*l,ncol=2))

for(n in 1:N){
  phis[((n-1)*(l)+1):((n)*l),1]=eval(parse(text=(paste0("phi.",n))))
  phis[((n-1)*(l)+1):((n)*l),2]=n
}

names(phis)=c("phi","County_Index")

ggplot(data = phis,aes(x= County_Index, y = phi,group=County_Index))+
  geom_boxplot(outlier.shape = NA)+scale_y_continuous(name=bquote(phi))+
  scale_x_discrete(limits = c(1,5,10,15,20,25,30,35,40,45,50,55,58))+
  ggtitle("County Spatial Effects")+theme(plot.title = element_text(hjust = 0.5))+
  geom_hline(yintercept=0, linetype="solid", color = "red")+
  annotate("text", x = 18, y = -1.5, label = cnames[18])+
  annotate("segment", color="red", x=19, xend = 18, y=-1.45, 
           yend=quantile(phi.18,0.25), arrow=arrow(length=unit(0.2,"cm")))+
  
  annotate("text", x = 20, y = 2.1, label = cnames[19])+
  annotate("segment", color="red", x=20, xend = 19, y=2, 
           yend=quantile(phi.19,0.75), arrow=arrow(length=unit(0.2,"cm")))+
  
  annotate("text", x = 37, y = 1.5, label = cnames[38])+
  annotate("segment", color="red", x=37, xend = 38, y=1.45, 
           yend=quantile(phi.38,0.75), arrow=arrow(length=unit(0.2,"cm")))+
  
   annotate("text", x = 32, y = -0.55, label = cnames[37])+
  annotate("segment", color="red", x=31, xend = 37, y=-0.5, 
           yend=quantile(phi.37,0.25), arrow=arrow(length=unit(0.2,"cm")))+
  
  annotate("text", x = 33, y = 1.75, label = cnames[36])+
  annotate("segment", color="red", x=33, xend = 36, y=1.7, 
           yend=quantile(phi.36,0.75), arrow=arrow(length=unit(0.2,"cm")))+
  
        annotate("text", x = 12, y = 1.1, label = cnames[15])+
  annotate("segment", color="red", x=12, xend = 15, y=1.1, 
           yend=quantile(phi.15,0.75), arrow=arrow(length=unit(0.2,"cm")))

phi_means=aggregate(.~County_Index,data=phis,mean)


ggplot() +
  annotation_spatial(shapeanddata) +
  ggtitle(bquote("Posterior Mean of "~phi~" s"))+
  theme(plot.title = element_text(hjust = 0.5))+
  layer_spatial(shapeanddata, aes(fill = (phi_means$phi)))+
  theme(legend.title= element_blank())+
  labs(fill =bquote(phi) )+scale_fill_viridis(limits = c(-2,2),direction=-1)+
  theme(legend.position = "none")

```

## $\rho$, Coefficient Structured vs Unstructured Errors
```{r, warning=FALSE,message=FALSE}

rhos=as.data.frame(matrix(nrow=M,ncol=2))
l=length(rho.1)
for(n in 1:ITER){
  rhos[((n-1)*(l)+1):((n)*l),1]=eval(parse(text=(paste0("rho.",n))))
  rhos[((n-1)*(l)+1):((n)*l),2]=n
}

names(rhos)=c("rho","Biweeks")

ggplot(data = rhos,aes(x= Biweeks, y = rho,group=Biweeks))+
  geom_boxplot(outlier.shape = NA)+scale_y_continuous(name=bquote(rho[t]))+
  geom_hline(yintercept=0.5, linetype="solid", color = "red") +
  scale_x_continuous(name="Time",
  labels=c('1','5','10','15','20','25','30','35','40','45','50',
'55','60','65','70','75','77'),breaks=c(1,5,10,15,20,25,30,35,40,45,50,
55,60,65,70,75,77))+ggtitle("Spat. Structured vs Uncorrelated Error")+
  theme(plot.title = element_text(hjust = 0.5))
```

## $\sigma$ Overall standard deviation

```{r, warning=FALSE,message=FALSE}

sigmas=as.data.frame(matrix(nrow=M,ncol=2))
l=length(rho.1)
for(n in 1:ITER){
  sigmas[((n-1)*(l)+1):((n)*l),1]=eval(parse(text=(paste0("sigma.",n))))
  sigmas[((n-1)*(l)+1):((n)*l),2]=n
}

names(sigmas)=c("sigma","Biweeks")

ggplot(data = sigmas,aes(x= Biweeks, y = sigma,group=Biweeks))+
  geom_boxplot(outlier.shape = NA)+scale_y_continuous(name=bquote(sigma[t]))+
  ggtitle("Overall Standard Deviation of Errors")+
  theme(plot.title = element_text(hjust = 0.5))+
  scale_x_continuous(name="Time",
                     labels=c('1','5','10','15','20','25','30','35',
                              '40','45','50','55','60','65','70','75','77'),breaks=c(1,5,10,15,20,25,30,35,
                                                                                     40,45,50,55,60,65,70,75,77))
```

## Convolution

Recall that convolution is the aggregated areal and random effects associated with the counties. It allows us to look into 

```{r, warning=FALSE,message=FALSE}



cv=matrix(nrow=ITER,ncol=N)
for(n in 1:N){
for(t in 1:ITER){
cv[t,n]=mean(
  ((sqrt(eval(parse(text=(paste0("rho.",t))))/scaling_factor))*eval(parse(text=(paste0("phi.",n))))+
  (sqrt(1-eval(parse(text=(paste0("rho.",t))))))*eval(parse(text=(paste0("theta.",t,".",n)))))*
  eval(parse(text=(paste0("sigma.",t))))
)
}}

convolution1=as.data.frame(matrix(nrow=6*ITER,ncol=3))
the6counties=c(19,37,38,18,15,36)
for(i in 1:6){
  convolution1[((i-1)*ITER+1) :((i-1)*ITER+ITER),1]=cv[,the6counties[i]]
  convolution1[((i-1)*ITER+1) :((i-1)*ITER+ITER),2]=rep(cnames[the6counties[i]],ITER)
  convolution1[((i-1)*ITER+1) :((i-1)*ITER+ITER),3]=c(1:ITER)
}
names(convolution1)=c("Convolution","County_Names","Biweeks")

ggplot(data = convolution1,aes(x= Biweeks, y = Convolution,group=County_Names))+
  geom_line(aes(color=County_Names,linetype=County_Names))+
  geom_hline(yintercept=0, linetype="solid", color = "red")+labs(y=bquote(sigma[t]*nu[i*","*t]))+
  annotate("text", x =which.min(convolution1[convolution1[,2]==cnames[38],1])-.25, 
           y =min(convolution1[convolution1[,2]==cnames[38],1])-.05, label = cnames[38])+
  annotate("text", x = which.max(convolution1[convolution1[,2]==cnames[19],1])-.75, 
           y =max(convolution1[convolution1[,2]==cnames[19],1])+.05, label = cnames[19])+
  annotate("text", x = which.min(convolution1[convolution1[,2]==cnames[37],1])-.25, 
           y =min(convolution1[convolution1[,2]==cnames[37],1])-.05, label = cnames[37])+
  annotate("text", x = which.max(convolution1[convolution1[,2]==cnames[36],1])-.25, 
           y =max(convolution1[convolution1[,2]==cnames[36],1])+.05, label = cnames[36])+
  annotate("text", x = which.min(convolution1[convolution1[,2]==cnames[15],1])-.25, 
           y =min(convolution1[convolution1[,2]==cnames[15],1])-.05, label = cnames[15])+
  annotate("text", x = which.max(convolution1[convolution1[,2]==cnames[18],1])-.25, 
           y =max(convolution1[convolution1[,2]==cnames[18],1])-.05, label = cnames[18])+
scale_x_continuous(name="Time",labels=c('1','5','10','15','20','25','30','35','40','45','50','55','60','65','70','75','77'),
breaks=c(1,5,10,15,20,25,30,35,40,45,50,55,60,65,70,75,77))+ggtitle("Convolution")+theme(plot.title = element_text(hjust = 0.5))

```

# References


